An in-depth study of Explainability and Adversarial Robustness for RNNs

### 2019-12-07 ex3/paper/paper.tex
Abbreviated IAT

We implement a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use \begin{itemize*}
\item source port
\item destination port
\item protocol identifier
\item packet sizes
\item interarrival times of packets
\item packet directions (if packet goes there or back)
\item all \gls{tcp} flags (0 if flow is not \gls{tcp})
\end{itemize*}. We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behaviour \cite{bachl_walling_2019}.  Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
We use Z-score normalization and a train/test split of 2:1.

@@

\section{Adversarial attacks}
\label{sec:adv}
We now investigate whether known \gls{aml} methods can be used for generating adversarial flows, i.e. minimally different attack flows which are classified as benign.
In our scenario, most features such as, e.g., \gls{tcp} flags cannot be easily manipulated because their manipulation might violate the \gls{tcp} and thus cause communication to fail. Hence, we experiment whether by manipulating the interarrival times and packet lengths of the first packets, the classifier might be fooled.
Our scenario is quite different from how generating adversarial samples usually works: We have significantly less features than in, for example, an image. Furthermore, only two out of those few features can be manipulated. And even these features cannot be manipulated freely: Besides the constraints that packet size and interarrival time cannot become negative there are also more problem-specific constraints:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Only packets can be manipulated which are transmitted by the attacker. This means the attacker can only control one direction (except for botnets where both sides are attackers).
\item Packets must not be smaller than initially, as otherwise less information could be transmitted or packets might be rendered syntactically invalid.
\item Interarrival times must not decrease, as otherwise the physical speed of data transmission can be violated in some cases. Since an in-depth analysis of when the reduction of interarrival times is possible, would come with substantial effort, we generally disallowed a reduction of this feature.
\end{itemize}

@@

The method minimizes the adversarial sample's distance so that the network's decision is just between attack and normal traffic. In the present context, we need to make sure that the classifier's prediction would actually be normal traffic, even though a certain level of noise will be added to interarrival times due to the network between attacker and victim. Instead of the above equation \ref{eq:carliniWagner} we therefore used the optimization objective
\begin{equation}
d(X,\tilde X) + \epsilon  \max(Z(\tilde X), -0.2),
\end{equation}
i.e. we require a slight level of confidence that our adversarial samples are not classified as attack traffic. A logit value of $-0.2$ corresponds to a confidence of 55\% for the sample to be benign.

@@

To this end, we generated adversarial flows with the \gls{cw} with $\ell_2$ distance and a tradeoff of 0.5. \autoref{adv_plots} depicts how the method manipulates flows for some exemplary attack categories. Depicted is the mean of both unmodified flows and adversarial flows  of samples of individual attack families considering both interarrival time and packet length.

@@

\subsubsection{Results}
\begin{figure*}
\includegraphics[width=.49\textwidth]{../plots/importance/feat_imp_flow_2017.pdf}\hspace{.02\textwidth}
\includegraphics[width=.49\textwidth]{../plots/importance/feat_imp_flow_2015.pdf}
\caption{Feature importance metrics considering the final flow prediction for CIC-IDS-2017 (left side) and UNSW-NB15 (right side).}
\label{fig:feat_imp}
\end{figure*}
Table~\ref{fig:featRandom} and \autoref{fig:feat_imp} show the results, which match to a large extent with domain-specific expectations for classifying network flows.
In particular, rarely used \gls{tcp} flags like NS or URG are unimportant for the classifier. On the other hand, destination port and protocol are essential for characterizing flows by hinting at the type of traffic. Interarrival time and packet length are important for estimating the amount of transferred information and several flags hint at certain attacks like \gls{dos} attacks.

@@

Next, we want to investigate if our classifier just memorized its inputs. The most interesting packet features that can be used to visualize this in our use case are the inter-arrival time (IAT) between packets in a flows and the packet length. To depict how the model comes to a decision, we determined the total minimum and maximum for these features that ever occur for all flows. For a given flow $X$ we then varied feature values at each time step between minimum and maximum and plot how the confidence changes when the feature varies (\autoref{fig:pred_plots2}) depict the averaged predictions for different attack families. It seems clear that the classifier didn't just memorize the inputs but actually learned them since there is always a smooth decrease of confidence if we change the feature values. Instead, when memorizing, we would expect a sharp decline of confidence to occur. Another interesting observation is that after the first couple of packets the classifier is certain about its decision and changing features doesn't change its confidence any longer. This allows the conclusion that if an attacker is able to change the first few packets, she could possibly change the decision of the classifier. We will further investigate this question in section \autoref{sec:adv}.

@@

Looking at the plots of the features that vary over time (\autoref{fig:plot_features}) we see that certain attack types have a characteristic pattern in which they send packets by which they are easily recognizable. We want to determine how the pattern of packet sizes and interarrival times is significant for the classification. \autoref{fig:plot_features} clearly shows that some attack types have very characteristic patterns of packet sizes and interarrival times.

@@

\subsection{Reducing features}
The most obvious defence, which is only possible because of the nature of the dataset, is to simply leave out features which the attacker can manipulate. For this, we try two different approaches:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Leaving out all features that are manipulable: packet size and interarrival time
\item Leaving out manipulable features when the attacker can manipulate them: only in the direction from the attacker to the victim (This, however does not prevent adversarial samples for botnets, for which both sides are malicious)
\end{itemize}

Both approaches lead to complete resistance to adversarial samples (as we left out all manipulable features), except for botnets, which can still operate when only leaving out manipulable features in one direction. The results in \autoref{tab:performance_results_no_manipulable} show that -- surprisingly -- there is only a small difference in classification performance when only computing the accuracy for the last packet of each flow (flow accuracy). However, when looking at the packet accuracy, which considers all packets of all flows, there is a significant difference. Thus, apparently the interarrival time and packet size are especially important for determining whether a flow is malicious in the first packets of a flow.

@@

\begin{table*}
\caption{Comparing performance with all features, only safe features (interarrival time and packet size removed) and safe features in one direction (interarrival time and packet size removed for packets coming from source). \textit{packets} means that we consider the neural network output of all packets of the flow while \textit{flows} means that we only consider the last packet. At the last packet the classifier is usually already a lot more confident than at the first packets of a flow and thus metrics are always better in this case.}  \label{tab:performance_results_no_manipulable}
\newcommand{\cmidrulespace}{6pt}
\centering
\begin{tabular}{l l l l l l l} \toprule
& \multicolumn{2}{l}{all} & \multicolumn{2}{l}{safe features} & \multicolumn{2}{l}{safe features one direction} \\
\cmidrule(r){2-3} \cmidrule(lr){4-5} \cmidrule(l){6-7}
& packets & flows & packets & flows & packets & flows \\
\midrule
Accuracy & 0.9857054 & 0.99472587 & 0.96678023 & 0.99337596 & 0.98056073 & 0.99353386 \\
Precision & 0.94963342 & 0.9894422 & 0.87175275 & 0.98798596 & 0.91359787 & 0.98994595 \\
Recall & 0.96735895 & 0.98972051 & 0.94374958 & 0.98581328 & 0.97836672 & 0.9844478 \\
F1 & 0.95841423 & 0.98958134 & 0.90632359 & 0.98689842 & 0.94487366 & 0.98718922 \\
Youden & 0.95682951 & 0.98614231 & 0.91525628 & 0.98175163 & 0.95937772 & 0.9810602 \\
\bottomrule
\end{tabular}
\end{table*}

>>>>

We implement a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use \begin{itemize*}
\item source port
\item destination port
\item protocol identifier
\item packet sizes
\item \glspl{iat} of packets
\item packet directions (if packet goes there or back)
\item all \gls{tcp} flags (0 if flow is not \gls{tcp})
\end{itemize*}. We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behaviour \cite{bachl_walling_2019}.  Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
We use Z-score normalization and a train/test split of 2:1.

@@

\section{Adversarial attacks}
\label{sec:adv}
We now investigate whether known \gls{aml} methods can be used for generating adversarial flows, i.e. minimally different attack flows which are classified as benign.
In our scenario, most features such as, e.g., \gls{tcp} flags cannot be easily manipulated because their manipulation might violate the \gls{tcp} and thus cause communication to fail. Hence, we experiment whether by manipulating the \glspl{iat} and packet lengths of the first packets, the classifier might be fooled.
Our scenario is quite different from how generating adversarial samples usually works: We have significantly less features than in, for example, an image. Furthermore, only two out of those few features can be manipulated. And even these features cannot be manipulated freely: Besides the constraints that packet size and \gls{iat} cannot become negative there are also more problem-specific constraints:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Only packets can be manipulated which are transmitted by the attacker. This means the attacker can only control one direction (except for botnets where both sides are attackers).
\item Packets must not be smaller than initially, as otherwise less information could be transmitted or packets might be rendered syntactically invalid.
\item \glspl{iat} must not decrease, as otherwise the physical speed of data transmission can be violated in some cases. Since an in-depth analysis of when the reduction of \glspl{iat} is possible, would come with substantial effort, we generally disallowed a reduction of this feature.
\end{itemize}

@@

The method minimizes the adversarial sample's distance so that the network's decision is just between attack and normal traffic. In the present context, we need to make sure that the classifier's prediction would actually be normal traffic, even though a certain level of noise will be added to \glspl{iat} due to the network between attacker and victim. Instead of the above equation \ref{eq:carliniWagner} we therefore used the optimization objective
\begin{equation}
d(X,\tilde X) + \epsilon  \max(Z(\tilde X), -0.2),
\end{equation}
i.e. we require a slight level of confidence that our adversarial samples are not classified as attack traffic. A logit value of $-0.2$ corresponds to a confidence of 55\% for the sample to be benign.

@@

To this end, we generated adversarial flows with the \gls{cw} with $\ell_2$ distance and a tradeoff of 0.5. \autoref{adv_plots} depicts how the method manipulates flows for some exemplary attack categories. Depicted is the mean of both unmodified flows and adversarial flows  of samples of individual attack families considering both \gls{iat} and packet length.

@@

\subsubsection{Results}
\begin{figure*}
\includegraphics[width=.49\textwidth]{../plots/importance/feat_imp_flow_2017.pdf}\hspace{.02\textwidth}
\includegraphics[width=.49\textwidth]{../plots/importance/feat_imp_flow_2015.pdf}
\caption{Feature importance metrics considering the final flow prediction for CIC-IDS-2017 (left side) and UNSW-NB15 (right side).}
\label{fig:feat_imp}
\end{figure*}
Table~\ref{fig:featRandom} and \autoref{fig:feat_imp} show the results, which match to a large extent with domain-specific expectations for classifying network flows.
In particular, rarely used \gls{tcp} flags like NS or URG are unimportant for the classifier. On the other hand, destination port and protocol are essential for characterizing flows by hinting at the type of traffic. \gls{iat} and packet length are important for estimating the amount of transferred information and several flags hint at certain attacks like \gls{dos} attacks.

@@

Next, we want to investigate if our classifier just memorized its inputs. The most interesting packet features that can be used to visualize this in our use case are the \gls{iat} between packets in a flows and the packet length. To depict how the model comes to a decision, we determined the total minimum and maximum for these features that ever occur for all flows. For a given flow $X$ we then varied feature values at each time step between minimum and maximum and plot how the confidence changes when the feature varies (\autoref{fig:pred_plots2}) depict the averaged predictions for different attack families. It seems clear that the classifier didn't just memorize the inputs but actually learned them since there is always a smooth decrease of confidence if we change the feature values. Instead, when memorizing, we would expect a sharp decline of confidence to occur. Another interesting observation is that after the first couple of packets the classifier is certain about its decision and changing features doesn't change its confidence any longer. This allows the conclusion that if an attacker is able to change the first few packets, she could possibly change the decision of the classifier. We will further investigate this question in section \autoref{sec:adv}.

@@

Looking at the plots of the features that vary over time (\autoref{fig:plot_features}) we see that certain attack types have a characteristic pattern in which they send packets by which they are easily recognizable. We want to determine how the pattern of packet sizes and \glspl{iat} is significant for the classification. \autoref{fig:plot_features} clearly shows that some attack types have very characteristic patterns of packet sizes and \glspl{iat}.

@@

\subsection{Reducing features}
The most obvious defence, which is only possible because of the nature of the dataset, is to simply leave out features which the attacker can manipulate. For this, we try two different approaches:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Leaving out all features that are manipulable: packet size and \gls{iat}
\item Leaving out manipulable features when the attacker can manipulate them: only in the direction from the attacker to the victim (This, however does not prevent adversarial samples for botnets, for which both sides are malicious)
\end{itemize}

Both approaches lead to complete resistance to adversarial samples (as we left out all manipulable features), except for botnets, which can still operate when only leaving out manipulable features in one direction. The results in \autoref{tab:performance_results_no_manipulable} show that -- surprisingly -- there is only a small difference in classification performance when only computing the accuracy for the last packet of each flow (flow accuracy). However, when looking at the packet accuracy, which considers all packets of all flows, there is a significant difference. Thus, apparently the \gls{iat} and packet size are especially important for determining whether a flow is malicious in the first packets of a flow.

@@

\begin{table*}
\caption{Comparing performance with all features, only safe features (\gls{iat} and packet size removed) and safe features in one direction (\gls{iat} and packet size removed for packets coming from source). \textit{packets} means that we consider the neural network output of all packets of the flow while \textit{flows} means that we only consider the last packet. At the last packet the classifier is usually already a lot more confident than at the first packets of a flow and thus metrics are always better in this case.}  \label{tab:performance_results_no_manipulable}
\newcommand{\cmidrulespace}{6pt}
\centering
\begin{tabular}{l l l l l l l} \toprule
& \multicolumn{2}{l}{all} & \multicolumn{2}{l}{safe features} & \multicolumn{2}{l}{safe features one direction} \\
\cmidrule(r){2-3} \cmidrule(lr){4-5} \cmidrule(l){6-7}
& packets & flows & packets & flows & packets & flows \\
\midrule
Accuracy & 0.9857054 & 0.99472587 & 0.96678023 & 0.99337596 & 0.98056073 & 0.99353386 \\
Precision & 0.94963342 & 0.9894422 & 0.87175275 & 0.98798596 & 0.91359787 & 0.98994595 \\
Recall & 0.96735895 & 0.98972051 & 0.94374958 & 0.98581328 & 0.97836672 & 0.9844478 \\
F1 & 0.95841423 & 0.98958134 & 0.90632359 & 0.98689842 & 0.94487366 & 0.98718922 \\
Youden & 0.95682951 & 0.98614231 & 0.91525628 & 0.98175163 & 0.95937772 & 0.9810602 \\
\bottomrule
\end{tabular}
\end{table*}
